---
title: "DV Assignment 2, Part 1: Amstelveen schools"
author: "Luna van Wonderen"
date: "`r Sys.Date()`"
output: html_document
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

## Research question

**What is your research question?**

Which schools in Amstelveen are on track for reading and maths?

**Is it a question of analytics, inference or prediction?**

Analytics- the goal is to describe and compare school performance across a municipality and identify which schools are on track for maths and reading.

**Who is the target audience for your final visualization, and how does that shape your choices?**

The target audience could be a municipal education officer or a school board.

## Exploratory process

**Describe how you explored the data. Include at least one exploratory visualization below, and a short note on what it shows, and what you learned from it about your research question.**
```{r prepare-data}

#Select relevant reference-level variables
count_cols <- c("REKENEN_LAGER1F", "REKENEN_1F", "REKENEN_1S",
                "LV_LAGER1F", "LV_1F", "LV_2F")

# value for DUO's "<5" (1-4 pupils): the midpoint
lt5_value <- 2.5

replace_lt5 <- function(x) {
  x <- as.character(x)
  as.numeric(if_else(x == "<5", as.character(lt5_value), x))
}

#Convert columns to numeric, and turn values like <5 to NA
#Clean reference-level counts
ref_levels <- referentieniveaus |>
  mutate(across(all_of(count_cols), replace_lt5)) |> #used to turn <5 into NA, now into 2.5 to prevent false 100$ scores? For small schools, this quite greatly influenced averages for example. 
  group_by(INSTELLINGSCODE) |>
  summarise(
    GEMEENTENAAM = first(GEMEENTENAAM),
    INSTELLINGSNAAM_VESTIGING = first(INSTELLINGSNAAM_VESTIGING),
    across(all_of(count_cols), ~ sum(.x, na.rm = TRUE)), .groups = "drop") |>
  mutate(
    #total number of students for maths and for reading
    rekenen_n = REKENEN_LAGER1F + REKENEN_1F + REKENEN_1S,
    reading_n = LV_LAGER1F + LV_1F + LV_2F,
    #percentage of students for each level
    math_1F = 100 * (REKENEN_1F + REKENEN_1S) / rekenen_n,
    math_1S = 100 * REKENEN_1S / rekenen_n,
    reading_1F = 100 * (LV_1F + LV_2F) / reading_n,
    reading_2F = 100 * LV_2F / reading_n
    #DUO only count each pupil once, so to get the totals for reaching 1F in both reading and math, you must also count         the pupils that reached 1S and 2F
  ) |>
  filter(rekenen_n > 0, reading_n > 0)
    #filter out schools where no pupils are counted for maths and reading, this might not occur for Amstelveen, but is          necessary for calculating averages.

#calculate Schoolweging
school_weights <- schoolweging |>
  mutate(INSTELLINGSCODE = str_remove(OVT, "\\|.*$")) |>
  group_by(INSTELLINGSCODE) |>
  summarise(schoolweging = weighted.mean(schoolweging, w = aantal_leerlingen),
    .groups = "drop")

#Combine reference-level results with schoolweging
school_data <- left_join(ref_levels, school_weights, by = "INSTELLINGSCODE")

#Amstelveen Gemeente
amstelveen_data <- filter(school_data, GEMEENTENAAM == "Amstelveen")

#Final data
amstelveen_data |>
  select(INSTELLINGSNAAM_VESTIGING, schoolweging,
          math_1F, math_1S, reading_1F, reading_2F)

```
## Exploratory plots
### Calculate National Average
```{r national-average}
national <- school_data |>
  summarise(
    # weighted.mean is weighted by the number of pupils. A school with 80
    # pupils counts more than a school with 10.
    math_1S = weighted.mean(math_1S, w = rekenen_n),
    reading_2F = weighted.mean(reading_2F, w = reading_n)
  )
```
### Check <5
In a first version of the scatterplot, where "<5" was treated as 0
(as in Assignment 1), 8 Amstelveen schools appeared at exactly 100%
reading. The check below shows why.
```{r check-lt5}
# 8 schools at 100% reading, this could be due to the <5 being
# transformed into 0 for 1F, due to which 'all' pupils appeared to have
# reached 2F.
# check this:
referentieniveaus |>
  filter(GEMEENTENAAM == "Amstelveen") |>
  select(INSTELLINGSNAAM_VESTIGING, LV_LAGER1F, LV_1F, LV_2F)
# appears that Parcival, het Palet, Kindercampus King, de Akker,
# de Triangel, Rembrandt, School of Understanding and Elif Amstelveen
# possibly do not really score 100%. Based on this, decided to change the
# schools that had <5 pupils to 2.5 instead of the original 0 as to not
# report untrue 100% scores.
```

### Create quadrant names
```{r quadrants}
amstelveen_data <- amstelveen_data |>
  mutate(
    maths_above = math_1S >= national$math_1S,
    reading_above = reading_2F >= national$reading_2F,
    quadrant = case_when(
      maths_above & reading_above ~ "Above average in both",
      !maths_above & !reading_above ~ "Below average in both",
      TRUE ~ "Above average in one subject"
    )
  )
```

### Exploratory plot
```{r exploratory-plot, fig.width = 10, fig.height = 7}
# one-row data frames to label the two national average lines
reading_label <- tibble(
  x = national$reading_2F,
  y = Inf,
  label = "National average reading"
)
maths_label <- tibble(
  x = Inf,
  y = national$math_1S,
  label = "National average maths"
)
ggplot(data = amstelveen_data,
       mapping = aes(x = reading_2F, y = math_1S)) +
  geom_vline(xintercept = national$reading_2F,
             linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = national$math_1S,
             linetype = "dashed", color = "grey50") +
  geom_label(data = reading_label, aes(x = x, y = y, label = label),
             hjust = 0, vjust = 1, size = 3, fontface = "italic",
             color = "grey40") +
  geom_label(data = maths_label, aes(x = x, y = y, label = label),
             hjust = 1, vjust = 0, size = 3, fontface = "italic",
             color = "grey40") +
  geom_point(aes(color = quadrant), size = 3) +
  geom_text_repel(aes(label = INSTELLINGSNAAM_VESTIGING), size = 3,
                  box.padding = 0.5, seed = 1, segment.color = NA) +
  scale_color_manual(
    values = c(
      # colours also visible for colour-blind readers!
      "Above average in both" = "#009E73",
      "Above average in one subject" = "#E69F00",
      "Below average in both" = "#B2182B"
    ),
    breaks = c(
      "Above average in both",
      "Above average in one subject",
      "Below average in both"
    )
  ) +
  labs(
    title = "Amstelveen schools compared to the national average",
    x = "% of pupils reaching reading level 2F",
    y = "% of pupils reaching maths level 1S",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Before submitting:** run the lintr check on this file, knit it, and push your work to GitHub.

#run lint(rstudioapi::getSourceEditorContext()$path)
